---
title: "Suivi_datasets"
author: "Kilian BARREIRO"
date: "2025-04-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("suivi_datasets_extra.R")
```


List of references
```{r}
updated_list_of_ref_biogeography_FP <- read_csv("~/Papers/French Polynesia Biodiversity Assessment/updated_list_of_ref_biogeography_FP.csv") %>% 
  select(
    Year, Title, Authors, Habitat, Source_type, Issue_Type, Data_merged,
    Geospatial_data_available, Geospatial_level, Repository, Listed_by, Notes
  ) %>% 
  arrange(desc(Year))

write_csv(updated_list_of_ref_biogeography_FP, "~/Papers/French Polynesia Biodiversity Assessment/updated_list_of_ref_biogeography_FP.csv")
```



## Expeditions
```{r}
# detect_date_formats(expeditionsMerged, col_name = "eventDate")
# detect_species_formats(expeditionsMerged, species_col = "species")
# identical(names(expeditionsMerged), names(filtered_data))
# all(map2_chr(expeditionsMerged, filtered_data, ~ class(.x)[1]) ==   map_chr(filtered_data, ~ class(.)[1]))

expeditionsMerged <- bind_rows(geodataset_BenthausExpedition, 
                               geodataset_Musorstom9Expedition, 
                               geodataset_TarasocExpedition, 
                               geodataset_PakaihiTeMoanaExpedition, 
                               geodataset_TuhaaPaeExpedition, 
                               geodataset_Rapa2002Expedition) %>%
# Dates
  mutate(
    eventDate_clean = str_trim(eventDate),
    eventDate_start = str_extract(eventDate_clean, "^[^/]+"),
    eventDate_start = str_replace_all(eventDate_start, "(\\d{4})-(\\d{1,2})-(\\d{1,2})", function(x) {
      parts <- str_match(x, "(\\d{4})-(\\d{1,2})-(\\d{1,2})")[,2:4]
      sprintf("%04d-%02d-%02d", as.integer(parts[1]), as.integer(parts[2]), as.integer(parts[3]))
    }),
    eventDate_converted = ymd(eventDate_start)
  ) %>%
  
#Species
  mutate(
    species = paste(genus, specificEpithet, sep = " "),
    nb_capital = str_count(species, "[A-Z]"),
    has_digit = str_detect(species, "\\d"),
    has_punct = str_detect(species, "[[:punct:]]")
  ) %>%
  
  filter(
    nb_capital == 1,               # Une majuscule au début
    !has_digit,                    # Pas de chiffres
    !has_punct                     # Pas de ponctuation
  ) %>%

  # Créer la colonne 'format' avant d'utiliser 'if_else'
  mutate(format = case_when(
    str_count(species, "\\S+") == 2 ~ "Deux mots",    # Si 2 mots
    str_count(species, "\\S+") == 3 ~ "Trois mots",    # Si 3 mots
    TRUE ~ "Autre"
  )) %>%

  # Appliquer le format pour conserver uniquement les espèces avec "Deux mots"
  mutate(species = if_else(format == "Deux mots", species, NA_character_)) %>%

  filter(!is.na(species),
         !is.na(eventDate)) %>% 

  mutate(
    day = as.numeric(format(eventDate_converted, "%d")),
    month = as.numeric(format(eventDate_converted, "%m")),
    year = as.numeric(format(eventDate_converted, "%Y"))
  ) %>%

# GBIF Formating
  mutate(gbifID = NA, publisher = paste(expedition, "(MNHN)"),
         references = NA, basisOfRecord = "HUMAN_OBSERVATION",
         individualCount = NA, occurrenceStatus = "PRESENT",
         identificationRemarks = recordedBy, kingdom = "Animalia",
         phylum = NA, subfamily = NA, genericName = NA, taxonomicStatus = NA,
         publishingCountry = "FR", elevation = NA, depth = NA, issue = NA,
         hasGeospatialIssues = NA, acceptedScientificName = scientificName,
         protocol = NA, level1Name = locality, iucnRedListCategory = NA) %>% 
  

  select(all_of(colnames(filtered_data))) %>% 

# Forcer les types des colonnes à ceux de filtered_data
  mutate(across(
    intersect(names(.), names(filtered_data)),
    ~ {
      target_class <- class(filtered_data[[cur_column()]])[1]
      match.fun(paste0("as.", target_class))(.)
    }
  )) %>% 
  
  filter(decimalLatitude >= -30, 
         decimalLatitude <= -5, decimalLongitude >= -155, 
         decimalLongitude <= -134)


# write.csv(expeditionsMerged, "ReadyToMerge_geodataset_EXPEDITIONS.csv", row.names = FALSE)
```



## FISHNET
```{r}
# To understand what has been done below for date formatting 
# species formatting
# detect_species_formats(BRANDL, species_col = "species")
# table(guess_formats(BRANDL_1$Date, c("ymd", "dmy", "mdy", "ydm")))
# columns order and format checking
# identical(names(Data_Sp_Obs), names(Original_Data_Sp_Obs))
# all(map2_chr(BRANDL, filtered_data, ~ class(.x)[1]) ==   map_chr(filtered_data, ~ class(.)[1]))
FISHNET <- read_csv("~/Papers/French Polynesia Biodiversity Assessment/Datasets to check in my GBIF almost only inspired dataset/FISHNET/SearchResults-20250417053816.csv") %>% 
  
  mutate(decimalLatitude = Latitude, 
    decimalLongitude = Longitude,
    species = ScientificName,
    year = YearCollected,
    month = MonthCollected,
    day = DayCollected) %>%
  
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude), !is.na(species),
         !str_detect(species, "[[:punct:]]"), str_count(species, "\\S+") >= 2,
         str_count(species, "[A-Z]") < 2, decimalLatitude >= -30, 
         decimalLatitude <= -5, decimalLongitude >= -155, decimalLongitude <= -134) %>%
  # We extract the two first words 
  mutate(species = str_replace(species, "^((\\S+)\\s+(\\S+)).*$", "\\1")) %>%
  
  mutate(
    gbifID = NA, publisher = "Fishnet", references = NA, 
    basisOfRecord = str_to_upper((str_replace_all(BasisOfRecord, "(?<=[a-z])(?=[A-Z])", "_"))),
    individualCount = IndividualCount, occurrenceStatus = "PRESENT",
    locality = Locality, identificationRemarks = Remarks, scientificName = species,
    kingdom = "Animalia", phylum = NA, class = NA, order = NA, family = Family, 
    subfamily = NA, genus = word(species, 1), genericName = NA,
    specificEpithet = word(species, 2), taxonRank = "SPECIES",
    taxonomicStatus = NA, publishingCountry = "FR", elevation = NA, 
    depth = VerbatimDepth, issue = NA, hasGeospatialIssues = NA, 
    acceptedScientificName = NA, protocol = PreparationType,
    level1Name = Island, iucnRedListCategory = NA) %>%
  
  select(all_of(colnames(filtered_data))) %>% 
  # Forcer les types des colonnes à ceux de Filtered_data_ori
  mutate(across(
    intersect(names(.), names(filtered_data)),
    ~ {
      target_class <- class(filtered_data[[cur_column()]])[1]
      match.fun(paste0("as.", target_class))(.)
    }
  ))


# write.csv(FISHNET, "ReadyToMerge_geodataset_FISHNET.csv", row.names = FALSE)

```


## ORP DATA SHARKS CLEMENTINE SEGUIGNE 2023  
https://doi.org/10.5061/dryad.z08kprrhm
5 espèces de requin manquantes et 31530 occurrences manquantes dans ma base de données. Citizen Science
GEOSPATIAL DATA
```{r}
library(tidyverse)
library(bit64)
# #Date format 
# formats_date_SEGUIGNE <- detect_date_formats(DataORP_obs_V3, col_name = "Date_observation")
# #Species Format Testing
# formats_esp_SEGUIGNE <- detect_species_formats(Data_Sp_Obs, species_col = "Species")

DataORP_obs_V3 <-  read_delim("~/Papers/French Polynesia Biodiversity Assessment/Datasets to check in my GBIF almost only inspired dataset/DRYAD/Sharks SEGUIGNE/DataORP_obs_V3.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE) %>% 
  mutate(Date_observation = as.Date(Date_observation, format = "%d/%m/%Y"),
         day = as.numeric(format(Date_observation, "%d")),
         month = as.numeric(Month),
         year = as.numeric(format(Date_observation, "%Y")),
         decimalLatitude = Lat,
         decimalLongitude = Long) %>%
  filter(decimalLatitude >= -30, decimalLatitude <= -5,
    decimalLongitude >= -155, decimalLongitude <= -134) 

Data_Sp_Obs <- read_delim("~/Papers/French Polynesia Biodiversity Assessment/Datasets to check in my GBIF almost only inspired dataset/DRYAD/Sharks SEGUIGNE/Data_Sp_Obs.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE) %>% 
  mutate(species = gsub("_", " ", Species),
         genus = str_split(species, " ", simplify = TRUE)[, 1],
         specificEpithet = str_split(species, " ", simplify = TRUE)[, 2]) %>%
  select(ID_fiche, ID_nombre, species, genus, specificEpithet)

ORPData <- left_join(DataORP_obs_V3, Data_Sp_Obs, by = "ID_fiche") %>%
  
  filter(!is.na(species)) %>% 
  filter(species != "Squalus") %>% 
  
  mutate(species = case_when(
    species %in% c("Dasyatis violacea","Raie violette") ~ "Pteroplatytrygon violacea",
    species == "Himantura fai" ~ "Pateobatis fai",
    species == "Requin Feroce" ~ "Odontaspis ferox",
    species == "Taeniura meyeni" ~ "Taeniurops meyeni",
    TRUE ~ species)) %>%
  
  mutate(
    gbifID = NA, publisher = "Polynesian Shark Observatory (ORP)",
    references = NA, basisOfRecord = "HUMAN_OBSERVATION", 
    individualCount = ID_nombre, occurrenceStatus = "PRESENT",
    locality = Name_Site, identificationRemarks = ID_observateur,
    scientificName = species, kingdom = "Animalia", phylum = NA,
    class = NA, order = NA, family = NA, subfamily = NA, genericName = NA,
    taxonRank = "SPECIES", taxonomicStatus = NA, publishingCountry = "FR",
    elevation = NA, depth = NA, issue = NA, hasGeospatialIssues = NA,
    acceptedScientificName = NA, protocol = NA, level1Name = Name_Site,
    iucnRedListCategory = NA
  ) %>% 
  
  select(all_of(names(filtered_data))) %>% 
#verification if columns formats are the same than filtered data
  mutate(across(
    intersect(names(.), names(filtered_data)),
    ~ {
      target_class <- class(filtered_data[[cur_column()]])[1]
      match.fun(paste0("as.", target_class))(.)
    }
  ))


#write.csv(ORPData, "ReadyToMerge_geodataset_SEGUIGNE_2020.csv", row.names = FALSE)
```



## BRANDL (2025) 
A seascape dichotomy in the role of small consumers for coral reef energy fluxes
https://doi.org/10.1002/ecy.70065 (papier)
https://doi.org/10.6084/m9.figshare.28400828.v1 (Data)
 insects, worms, mollusks, or small vertebrates
 GEOSPATIAL DATA (CRIOBE), 
BRANDL DATASET MERGED (0 new species, 778/778 new occurrences)
```{r}
# species formatting
# detect_species_formats(BRANDL, species_col = "species")
# To understand what has been done below for date formatting 
# table(guess_formats(BRANDL_1$Date, c("ymd", "dmy", "mdy", "ydm")))
# columns order and format checking
# identical(names(Data_Sp_Obs), names(Original_Data_Sp_Obs))
# all(map2_chr(BRANDL, filtered_data, ~ class(.x)[1]) ==   map_chr(BRANDL, ~ class(.)[1]))
BRANDL_1 <- readr::read_csv("~/Papers/French Polynesia Biodiversity Assessment/Datasets to check in my GBIF almost only inspired dataset/FIGSHARE/BRANDL 2025/moorea_communities_allyears.csv") %>%
  {list(
    part1 = .[1:1938, ] %>% mutate(Date = mdy(Date)),
    part2 = .[1939:nrow(.), ] %>% mutate(Date = dmy(Date))
  )} %>%
  {rbind(.$part1, .$part2)} %>%
  mutate(
    Date = as.Date(Date),
    day = as.numeric(format(Date, "%d")),
    month = as.numeric(format(Date, "%m")),
    year = as.numeric(format(Date, "%Y")),
    decimalLatitude = Lat, #round(Lat, 4),
    decimalLongitude = Long #round(Long, 4)
  ) %>%
  mutate(Species = str_extract(Species, "^\\S+")) %>% #extrait premier mot
  filter(!str_detect(Species, "[A-Z0-9]")) %>% #assure pas de Majuscule ou Chiffre
  mutate(species = paste(Genus, Species, sep = " ")) 

BRANDL_3 <- read_csv("~/Papers/French Polynesia Biodiversity Assessment/Datasets to check in my GBIF almost only inspired dataset/FIGSHARE/BRANDL 2025/crypto.sia.2019.csv") %>%
  mutate(
    Date = lubridate::dmy(Date),
    day = as.numeric(format(Date, "%d")),
    month = as.numeric(format(Date, "%m")),
    year = as.numeric(format(Date, "%Y")),
    decimalLatitude = Lat, #round(Lat, 4),
    decimalLongitude = Long #round(Long, 4)
  ) %>%
  mutate(Species = str_extract(Species, "^\\S+")) %>% #extrait premier mot
  filter(!str_detect(Species, "[A-Z0-9]")) %>% #assure pas de Majuscule ou Chiffre
  mutate(species = paste(Genus, Species, sep = " "))

# We compute what is in BRANDL_3 and not in BRANDL_1 to add it afterwards
columns_to_compare = c("day", "month", "year", "decimalLatitude", "decimalLongitude", "species")

diff_BRANDL3_vs_BRANDL1 <- anti_join(BRANDL_3, BRANDL_1, by = columns_to_compare) %>%
  mutate(across(
    intersect(names(.), names(BRANDL_1)),
    ~ suppressWarnings(as(., class(BRANDL_1[[cur_column()]])))
  )) %>% 
  select(-all_of(setdiff(colnames(BRANDL_3), colnames(BRANDL_1)))) %>% 
  select(all_of(colnames(BRANDL_1)))

# We put the two datasets together prior to general formatting
BRANDL <- bind_rows(BRANDL_1, diff_BRANDL3_vs_BRANDL1) %>% 
  distinct() %>%  
  mutate(
    gbifID = NA, publisher = "Figshare (CRIOBE) Brandl et al. 2025",
    references = NA, basisOfRecord = "HUMAN_OBSERVATION", 
    individualCount = NA, occurrenceStatus = "PRESENT",
    locality = Location, identificationRemarks = Full_ID, genus = Genus,
    scientificName = species, kingdom = "Animalia", phylum = NA,
    class = NA, order = NA, family = Family, subfamily = NA, genericName = NA,
    specificEpithet = Species,
    taxonRank = "SPECIES", taxonomicStatus = NA, publishingCountry = "FR",
    elevation = NA, depth = Depth, issue = NA, hasGeospatialIssues = NA,
    acceptedScientificName = NA, protocol = NA, level1Name = Habitat,
    iucnRedListCategory = NA) %>% 
  
  select(all_of(colnames(filtered_data))) %>% 
  filter(decimalLatitude >= -30, decimalLatitude <= -5,
    decimalLongitude >= -155, decimalLongitude <= -134) %>% 
  mutate(across(
    intersect(names(.), names(filtered_data)),
    ~ {
      target_class <- class(filtered_data[[cur_column()]])[1]
      match.fun(paste0("as.", target_class))(.)
    }
  ))
  

#write.csv(BRANDL, "ReadyToMerge_geodataset_BRANDL_2025.csv", row.names = FALSE)

```









WE MERGE EVERYTHING
Each dataset below have been formatted to fit the filtered data col names
```{r}
# Last checks :
# table(is.na(Annexe_Data_Merged$species))
# detect_species_formats(Annexe_Data_Merged, species_col = "species")
# identical(names(Annexe_Data_Merged), names(filtered_data))
# all(map2_chr(Annexe_Data_Merged, filtered_data, ~ class(.x)[1]) ==   map_chr(filtered_data, ~ class(.)[1]))

Annexe_Data_Merged <- bind_rows(FISHNET,
                                BRANDL,
                                expeditionsMerged,
                                ORPData) 

# write.csv(Annexe_Data_Merged, "Annexe_Merged_Geodataset.csv", row.names = FALSE)
```
